Relationship Between Common Stream Health Indicators

### 2021-12-12 products/manuscript/Manuscript.Rmd
Misc. Edits - CC

Data was acquired from a publicly available spreadsheet of water quality measurements collected by the Upper Oconee Watershed Network (UOWN) between 2003 and 2020. Data originated from three distinct watersheds in the Athens- Clarke County area: Middle Oconee (MIDO), North Oconee (NORO), and Oconee (BIOC). The data were analyzed first by using bivariate linear modeling of the variables conductivity, turbidity, nitrate (no3) concentration, E. coli cfu, and biodiversity indicator score (biological score). E. coli cfu and biological score were used as outcomes to specifically see if easy to measure physical/chemical properties could predict biological variables to assess water quality. Second, the MIDO data were analyzed using LASSO linear modeling to construct the best combinations of variables to predict E. coli cfu and biological score.

Results indicated that the only significant bivariate relationships that existed were between E. coli cfu and conductivity for the MIDO site only. LASSO modeling produced models that did not fit the data any better than the null model that was calculated for cross validation. This indicates that physical/chemical water quality measurements cannot predict biological measurements. This suggests that the citizen science approach to water quality monitoring should be re-evaluated for sampling robustness or that the effect of urbanization is spatially and temporally different and may not captured well without more in-depth sampling. Future studies should investigate how specific aspects of urbanization actually impact our indicator variables of water quality and stream health.

>>>>

Data was acquired from a publicly available spreadsheet of water quality measurements collected by the Upper Oconee Watershed Network (UOWN) between 2003 and 2020. Data originated from three distinct watersheds in the Athens- Clarke County area: Middle Oconee (MIDO), North Oconee (NORO), and Oconee (BICO). The data were analyzed first by using univariate linear modeling of the variables conductivity, turbidity, nitrate (no3) concentration, E. coli cfu, and biodiversity indicator score (biological score). E. coli cfu and biological score were used as outcomes to specifically see if easy to measure physical/chemical properties could predict biological variables to assess water quality. Second, the MIDO data were analyzed using LASSO linear modeling to construct the best combinations of variables to predict E. coli cfu and biological score.

Results indicated that the only significant univariate relationships that existed were between E. coli cfu and conductivity for the MIDO site only. LASSO modeling produced models that did not fit the data any better than the null model that was calculated for cross validation. This indicates that physical/chemical water quality measurements cannot predict biological measurements. This suggests that the citizen science approach to water quality monitoring should be re-evaluated for sampling robustness or that the effect of urbanization is spatially and temporally different and may not captured well without more in-depth sampling. Future studies should investigate how specific aspects of urbanization actually impact our indicator variables of water quality and stream health.
